---
title: "triact package for R: Analyzing the lying behavior of cows from accelerometer data"
author: "Michael Simmler, Stijn Brouwers"
date: "12.10.2022"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{triact}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(digits = 3)
```

\
This vignette contains executable examples of the intended use of the *triact* R package. With exceptions, the functionally is introduced using default parameters only. We advocate to also read the help page to learn about the many parameters available to tweak the behavior of methods in *triact*. Furthermore we detailed the background and rationale of the implemented analyses in the following publication:\

> Simmler M., Brouwers S., 2023. *triact package for R: Analyzing the lying behavior of cows from accelerometer data*, in preparation


## Setup

```{r}
library(triact)
```

\
As *triact* is intended to be used with data \>1 Hz sampling frequency it is advisable to adjust R's global option `digits.secs`, which controls the maximum number of digits to print when formatting time values in seconds (default is 0).\

```{r}
options(digits.secs = 3)
```

\
Via the global option `triact_table` the type of tables returned by *triact* can be specified.\

```{r}
options(triact_table = "data.frame") # alternatively set to "tibble" or "data.table"
```

## Getting help

All functionally of the *triact* package is documented on the help page of the Triact R6 class.\

```{r, eval=FALSE}
?Triact
```

\newpage

## Get and inspect the example data

The example data can be downloaded from <https://xyz.com>. The data includes four raw data files from triaxial accelerometers (MSR145, MSR Electronics, Switzerland) mounted to the left hind leg of dairy cows. The sampling frequency was 20 Hz. Each file represents one day of recording of one cow.\

```{r}

dir <-  "example_data" # adjust (path to the downloaded example data)

files <- list.files(dir)

print(files)

```

\
Inspecting one of the files reveals the file header and the semicolon-separated data starting after the line with `"*Data`". This is an example of how files ingested by *triact* might look. However, *triact* can handle any kind of delimiter-separated text files.\

```{r}

cat(paste(readLines(file.path(dir, files[1]), n = 30), collapse = "\n"))

```

## Importing data

The typical *triact* workflow starts by creating a new object of the `Triact` class.\

```{r}
my_triact <- Triact$new()
```

\
Accelerator data is then imported into the *triact* object. Using the `$load_files` method the data from raw data file can be imported. This method has a plethora of arguments, which allows adapting the behavior to read the specific format of the delimiter-separated data file. Important for correct analyses of the data are `id_substring` and `timeFwdUpRight_cols`. The first identifies the substring of the file names that uniquely identify the cows, while the latter maps the columns as found in the file to the *time*, *forward* acceleration, *up* acceleration, and *right* acceleration as understood by *triact*.\

```{r}
my_triact$load_files(input = dir,
                     id_substring = c(12, 14),
                     timeFwdUpRight_cols = c(1, 2 ,3, 4),
                     skip = "*DATA")
```

\
Imported raw data and added analyses can be accessed between any step of the workflow via the `$data` field.\

```{r}
head(my_triact$data)
```

\

```{r}
str(my_triact$data)
```

\newpage

## Add analyses

\
Once data is imported, analyses of standing versus lying, lying side, and activity proxies can be added to the *triact* object. The corresponding methods have prefix \$add\_

### Detecting standing and lying posture

The `$add_lying()` method performs the classification into lying and standing. The results are silently added to the data in the *triact* object as a column named *lying*.\

```{r}
my_triact$add_lying()
```

### Detecting lying laterality

The `$add_side()` method performs the determination of lying laterality (left/right). The results are silently added to the data in the *triact* object as a column named *side*. Crucial for correct determination of the lying side is the correct specification of the hind leg the accelerometer was mounted on (parameter `left_leg = TRUE` for left, or `FALSE` for right).\

```{r}
my_triact$add_side(left_leg = TRUE)
```

### Calculate proxies for physical activity

The `$add_activity()` method performs the calculation of proxies for the physical activity level of the cow(s), by default the L2 norm of the vector of the dynamic body acceleration (L2DBA) and the L2 norm of the jerk vector (L2Jerk).\

```{r}
my_triact$add_activity()
```

\
The data in the *triact* object now additionally contains the added analyses (last four columns).\

```{r}
str(my_triact$data)
```

\newpage

\
The *triact* package does not come with visualization capabilities. But the data can easily be accessed and plotted with base R or packages dedicated to graphics (e.g. ggplot2). The following example demonstrated how e.g. the data of the cow with ID 013 can be accessed and the lying behavior visualized.\

```{r}

cow_id = "013"

data_013 <- my_triact$data[my_triact$data$id == cow_id, ]
                           
plot(!lying ~ time, data = data_013,
     type = "l", ylab = "", yaxt = "n", bty = "n")

lines(ifelse(side == "R", 0, NA) ~ time, data = data_013, col = "orange")
lines(ifelse(side == "L", 0, NA) ~ time, data = data_013, col = "purple")

axis(2, at = c(0, 1), 
     labels = c("lying", "standing"), 
     las = 1, 
     lwd = 0)

legend(x = "center",          
       legend = c("right", "left"),
       col = c("orange", "purple"),       
       lwd = 1, bty = "n", title = "lying side") 


```

\

\newpage

## Summarize results

Using the methods with prefix \$summarize the analyses added to the *triact* object are summarized per time period, which represent either the standing/lying bouts or regular intervals, e.g. 1 h or 24 h.

### Summarize per bout

With `$summarize_bouts()` a summary is created for the individual lying and standing bouts, with duration, mean activity, and lying side (for a lying bout). In the output we see that the first bout per cow is not completely observed (`startTime` is missing) and therefor NAs are returned for measures such as duration. See parameter `calc_for_incomplete`in case you want to overwrite this behavior. 
\

```{r}

bouts_summary <- my_triact$summarize_bouts()

```

```{r}

head(bouts_summary)

```
\
If only the lying bouts are of interest, the `bout_type` parameter can be specified accordingly. 
\
```{r}

bouts_summary <- my_triact$summarize_bouts(bout_type = "lying")

```

```{r}
head(bouts_summary)
```
\newpage
### Summarize per regular intervals

With `$summarize_intervals()` the summary is obtained per regular intervals, by default per hour.\

```{r}
int_summary <- my_triact$summarize_intervals()
```
\
The NAs in the output are a result of incompletely observed intervals (first and last interval of each cow) and of dependencies of measures on information of incompletely observed bouts (can also affect second and further intervals). See parameter `calc_for_incomplete`in case you want to overwrite this behavior. The NaN on the other hand do not indicate missing information: For example, if the cow was not lying in the interval, the mean activity during lying is not 0, but cannot be calculated (thus NaN, "not a number"). 
\
```{r}

head(int_summary)

```

```{r}

str(int_summary)

```

\newpage
The intervals can be specified quite flexibly: In case of 30 min intervals and starting 10 min after the full hour we can specify `interval` and `lag_in_s` parameters accordingly.
\
```{r}

int_summary <- my_triact$summarize_intervals(interval = "30 min", 
                                             lag_in_s = 10 * 60)

```

```{r}

head(int_summary)

```

```{r}

str(int_summary)

```
\
\newpage
With `bouts == TRUE` we can request that additional the bouts within the intervals are summarized. For measures such as the number of lying bouts or mean lying bout duration, a weighted mean is calculated with the weights being the proportion of the individual bout overlapping with the respective interval. With `side = TRUE` we can additionally request a differentiation of all results by lying side.      
\
```{r}
int_summary <- my_triact$summarize_intervals(bouts = TRUE, 
                                             side = TRUE)
```

```{r}
str(int_summary)
```

\newpage
### Extract ...


\newpage
### Additional functionality 
















